Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-12 latex/main.tex
Updates from Overleaf
Add imputation analysis

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions
in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
when accounting for various differences in the data represented
(Document~\ref{sec:supplement_sc2ts_usher_parsimony}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

>>>>

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions
in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

\subsection{Imputation comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_imputation}
Both sc2ts and UShER support missing data in input alignments, and effectively
``hard call'' these values when the samples are integrated into the tree/ARG. 
In doing so, we impute the missing and ambiguous data for these samples.
We evaluate the imputation performance of sc2ts and UShER by comparing their
outputs on the Viridian alignments.

The total aligned dataset for 2,475,418 samples and 27,507 sites 
in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
Of the 162,838,461 missing data calls (Ns) in the alignments, sc2ts and UShER
disagreed in their imputed values for only 203,579 (0.125\%). 
Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed 435 (0.05\%) incorrectly (i.e., 
with a base that is not compatible with the
ambiguity code). 
UShER imputed only 7 calls (0.007\%) from this set incorrectly, likely 
due to special handling of ambiguous bases in the parsimony calculations.

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter notebook, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.

See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/example\_data\_processing.ipynb}{example\_data\_processing.ipynb}
for details.
